### 2020-04-03 thesis/content/chapter-graph-lta.tex
completed proof that 2-wl-gnns are at least as powerful as 1-wl

		\item there is an associative algebra of $\mathit{Pool}$,

@@

The reason for this is that the composition $\mathrm{\acs*{mlp}} \circ \mathit{Pool}$ cannot be guaranteed to be homomorphic to a weighted aggregation function $\mathcal{A}$ due to the universal approximation theorem.

>>>>

		\item there is an associative algebra of $\mathit{Pool}$ and

@@

The reason for this is that the composition $\mathrm{\acs*{mlp}} \circ \mathit{Pool}$ cannot be guaranteed to be homomorphic to a weighted aggregation function $\mathcal{A}$ due to the well-known universal approximation theorem~\cite{Hornik1991}.

### 2020-04-03 thesis/content/chapter-ltd.tex
completed proof that 2-wl-gnns are at least as powerful as 1-wl

		To address this issue we reduce both the number of colored edges as well as the number of neighbors of each edge.
		This is achieved by only considering the edges that are part of the so-called \textit{$r$-th power of a given graph $G$} where $r \in \mathbb{N}$ is the freely choosable \textit{neighborhood radius}.
		\begin{defn}
			The \textit{$r$-th power of a graph $G$} is defined as
			\begin{align*}
				G^r \coloneqq \left(\mathcal{V}_G, \left\{ e_{ij} \in \mathcal{V}_G^2\, |\, d_{\mathrm{SP}, G}(v_i, v_j) \leq r \right\}\right) % chktex 21
			\end{align*}
			where $d_{\mathrm{SP}, G}(v_i, v_j)$ is the length of the shortest path between $v_i$ and $v_j$ in $G$.
			The distance of a vertex $v_i \in \mathcal{V}_G$ to itself is defined as $d_{\mathrm{SP}, G}(v_i, v_i) \coloneqq 0$.
			Note that $G^1$ does not generally equal $G$ because $G^1$ has self-loop edges $e_{ii} \in \mathcal{E}_{G^1}$ at all vertices.
		\end{defn}
		For the neighborhood radius $r = 1$ only the self-loop edges ${\{ e_{ii} \}}_{v_i \in \mathcal{V}_G}$ and the edges $\mathcal{E}_{G}$ are considered;
		for $r > 1$ edges between indirectly connected vertices are considered as well.
		\Cref{fig:ltd:neighborhood-radius} illustrates this for varying neighborhood radii.
		Through the reduction of the considered edges, the neighbors of each $e_{ij} \in \mathcal{E}_{G^r}$ are in turn reduced to the common $r$-neighbors of $v_i$ and $v_j$, i.e.\ $\{ \ldblbrace e_{il}, e_{lj} \rdblbrace\, |\, v_l \in {\Gamma_{G^r}(v_i) \cap \Gamma_{G^r}(v_j)} \}$. % chktex 21
		\begin{figure}[t]
			\centering
			\includegraphics[width=0.9\linewidth]{gfx/ltd/neighborhood-radius.pdf}
			\caption{
				Illustration of the powers of the six-cycle graph for varying $r$.
				The self-loop edge at each of the vertices is not explicitly shown.
				For $r = 3$ all possible edges between the six vertices will be considered just as in the original 2-\acs{wl} algorithm.
			}\label{fig:ltd:neighborhood-radius}
		\end{figure}

@@

\begin{prop}
	2-\acs{wl}-\acsp{gnn} can simulate arbitrary \acp{gin}.
\end{prop}

>>>>

		To address this issue we reduce both the number of colored edges as well as the number of neighbors of each edge.
		This is achieved by only considering the edges that are part of the so-called \textit{$r$-th power of a given graph $G$} where $r \in \mathbb{N}$ is the freely choosable \textit{neighborhood radius}.
		\begin{defn}
			The \textit{$r$-th power of a graph $G$} is defined as
			\begin{align*}
				G^r \coloneqq \left(\mathcal{V}_G, \left\{ e_{ij} \in \mathcal{V}_G^2\, |\, d_{\mathrm{SP}, G}(v_i, v_j) \leq r \right\}\right) % chktex 21
			\end{align*}
			where $d_{\mathrm{SP}, G}(v_i, v_j)$ is the length of the shortest path between $v_i$ and $v_j$ in $G$.
			The distance of a vertex $v_i \in \mathcal{V}_G$ to itself is defined as $d_{\mathrm{SP}, G}(v_i, v_i) \coloneqq 0$.
			Note that $G^1$ does not generally equal $G$ because $G^1$ has self-loop edges $e_{ii} \in \mathcal{E}_{G^1}$ at all vertices.
		\end{defn}
		For the neighborhood radius $r = 1$ only the self-loop edges ${\{ e_{ii} \}}_{v_i \in \mathcal{V}_G}$ and the edges $\mathcal{E}_{G}$ are considered;
		for $r > 1$ edges between indirectly connected vertices are considered as well.
		\Cref{fig:ltd:neighborhood-radius} illustrates this for varying neighborhood radii.
		Through the reduction of the considered edges, the neighbors of each $e_{ij} \in \mathcal{E}_{G^r}$ are in turn reduced to the common $r$-neighbors of $v_i$ and $v_j$, i.e.\ $\{ \ldblbrace e_{il}, e_{lj} \rdblbrace\, |\, v_l \in {\Gamma_{G^r}(v_i) \cap \Gamma_{G^r}(v_j)} \}$. % chktex 21
		\begin{figure}[t]
			\centering
			\includegraphics[width=0.9\linewidth]{gfx/ltd/neighborhood-radius.pdf}
			\caption[Illustration of the powers of the six-cycle graph.]{
				Illustration of the powers of the six-cycle graph for varying $r$.
				The self-loop edge at each of the vertices is not explicitly shown.
				For $r = 3$ all possible edges between the six vertices will be considered just as in the original 2-\acs{wl} algorithm.
			}\label{fig:ltd:neighborhood-radius}
		\end{figure}

@@

\begin{defn}\label{defn:ltd:vert-conv}
	A \ac{gnn} $h_1: \mathcal{G} \to \mathcal{Y}$ uses \textit{weighted vertex neighborhood sums} if its convolutional layers can be described by
	\begin{align*}
		Z_{G}^{(t)}[v_i] = \mathrm{\acs*{mlp}}^{(t)}\left( w_{ii} Z_{G}^{(t-1)}[v_i] + \smashoperator[lr]{\sum_{v_j \in \Gamma_G(v_i)}} w_{ij} Z_{G}^{(t-1)}[v_l] \right)
		\text{.}
	\end{align*}
\end{defn}
\Cref{defn:ltd:vert-conv} includes \acp{gcn}~\cite{Kipf2017} where the \ac{mlp} only consists of a single layer and the weights are $w_{ij} = {\left(\left| \Gamma_G(v_i) \right| + 1\right)}^{-\frac{1}{2}}{\left(\left| \Gamma_G(v_j) \right| + 1\right)}^{-\frac{1}{2}}$ (see \cfullref{eq:related:gcn-wl1}).
\Acp{gin}~\cite{Xu2018} also trivially satisfy the definition (see \cfullref{eq:related:gin-layer}).
\begin{thm}\label{thm:ltd:wl2-simulation}
	For each \ac{gnn} $h_1$ that uses weighted vertex neighborhood sums, there is a 2-\acs{wl}-\acs{gnn} $h_2$ which simulates $h_1$, i.e.\ $\forall G \in \mathcal{G}: h_1(G) = h_2(G)$.
\end{thm}
\begin{proof}
	We prove the theorem by construction.
	Let $G \in \mathcal{G}$ be an arbitrary input graph with $n \coloneqq \left|\mathcal{V}_G\right|$ and $m \coloneqq n + \left|\mathcal{E}_G\right|$.
	By definition $h_1$ is a \ac{gnn} of the form $\mathit{Pool}_1(\mathit{Conv}_1(G))$ where $\mathit{Conv}_1$ is a stack of $T$ weighted vertex neighborhood sum convolutions ${\left\{ S^{(t)}: \mathbb{R}^{n \times d^{(t-1)}} \to \mathbb{R}^{n \times d^{(t)}} \right\}}_{t=1}^T$ with each corresponding $\mathrm{\acs*{mlp}}^{(t)}$ having $K$ layers. % chktex 21
	$\mathit{Pool}_1$ combines the vertex feature vectors produced by $\mathit{Conv}_1$.

	Let $h_2$ be a \ac{gnn} of the form $\mathit{Pool}_2(\mathit{Conv}_2(G))$ where $\mathit{Conv}_2$ is a stack of $(2 + K) T$ 2-\acs{wl} convolution layers ${\left\{ S^{(t, k)}: \mathbb{R}^{m \times d^{(t,k-1)}} \to \mathbb{R}^{m \times d^{(t,k)}} \right\}}_{(t, k) \in [T] \times [2+K]}$ with the neighborhood radius $r \coloneqq 1$. % chktex 21
	Thus $\mathit{Conv}_2$ produces a feature vector for each of the $m$ edges of $G^1$, i.e.\ one for each vertex and each edge of $G$.
	We denote the initial 2-\acs{wl} feature matrix with $Z^{(1,0)} \in \mathbb{R}^{m \times d^{(0,0)}}$.
	The layers ${\left\{ S^{(t, 2 + K)} \right\}}_{t=1}^T$ produce the feature matrices $Z^{(t, 2 + K)} = Z^{(t + 1, 0)}$ which are then fed as input into the successor layer $S^{(t + 1, 1)}$. % chktex 21
	Intuitively $\mathit{Conv}_2$ simulates each layer of $\mathit{Conv}_1$ via a stack of $2 + K$ 2-\acs{wl} convolution layers.
	This is illustrated in \cref{fig:ltd:wl2-simulation}.
	\begin{figure}[ht]
		\centering
		\includegraphics[width=\linewidth]{gfx/ltd/wl2-simulation.pdf}
		\caption[Illustration of a 2-\acs{wl}-\acs{gnn} architecture that simulates vertex neighborhood convolutions.]{
			Illustration of the correspondence between $\mathit{Conv}_1$ and $\mathit{Conv}_2$.
		}\label{fig:ltd:wl2-simulation}
	\end{figure}

	Let $\varphi: \mathbb{R}^{d^{(T, 2 + K)}} \to \mathbb{R}^{d^{(T)}} \cup \{ \texttt{nil} \}$ be a function which maps the final 2-\acs{wl} feature vectors produced by $\mathit{Conv}_2$ to the output space of $\mathit{Conv}_1$ or the constant \texttt{nil}.
	Let $\mathit{Pool}_2(Z_G^{(T,2+K)}) \coloneqq \mathit{Pool}_1(\ldblbrace z_{ij}\, |\, z_{ij} = \varphi(Z_G^{(T,2+K)}[e_{ij}]) \land e_{ij} \in \mathcal{E}_{G^1} \land z_{ij} \neq \texttt{nil} \rdblbrace)$.
	\Cref{thm:ltd:wl2-simulation} then follows if there is a function $\varphi$ s.t.\ $\forall v_i \in \mathcal{V}_G: \mathit{Conv}_1(G)[v_i] = \varphi(\mathit{Conv}_2(G)[e_{ii}])$ and $\forall e_{ij} \in \mathcal{E}_G: \varphi(\mathit{Conv}_2(G)[e_{ij}]) = \texttt{nil}$.
	To guarantee that there is such a function $\varphi$ we now inductively prove the following three invariants which have to hold for all $t \in \{ 0, \dots, T \}$:
	\begin{enumerate}[label={(P\arabic*)}]
		\item\label{inv:ltd:wl2-simulation:indicator}
			${Z_G^{(t,2+K)}[e_{ij}]}_1 = \mathbbm{1}[i = j]$, i.e.\ the first component of each 2-\acs{wl} feature vector allows $\varphi$ to decide whether it should be mapped to \texttt{nil}.
		\item\label{inv:ltd:wl2-simulation:feature}
			${Z_G^{(t,2+K)}[e_{ii}]}_{2, \dots, (d^{(t)} + 1)} = Z_G^{(t)}[v_{i}]$, i.e.\ the second to $(1 + d^{(t)})$-th components of each self-loop feature vector contain the corresponding convolved vertex feature vector at layer $t$.
		\item\label{inv:ltd:wl2-simulation:weight}
			${Z_G^{(t,2+K)}[e_{ij}]}_{d^{(t)} + 2} = w_{ij}$, i.e.\ the weights for the vertex neighborhood sums are encoded in the edge and self-loop feature vectors.
	For $t = 0$ all three invariants hold by \cref{defn:ltd:wl2-conv-init}:
		\forall v_i \in \mathcal{V}_G: Z_G^{(1, 0)}[e_{ii}] \coloneqq (1) \oplus x[v_i] \oplus (w_{ii})
		\text{ and }
		\forall e_{ij} \in \mathcal{E}_G: Z_G^{(1, 0)}[e_{ij}] \coloneqq (0) \oplus \mathbf{0} \oplus (w_{ij})
	Assuming the invariants hold for $t - 1$ we now show that they also hold for $t$.
	As illustrated in \cref{fig:ltd:wl2-simulation} the layers $S^{(t, 1)}$ and $S^{(t, 2)}$ should compute the weighted vertex neighborhood sums
		{Z^{(t, 2)}[e_{ii}]}_{2, \dots, (1+d^{(t-1)})} = w_{ii} {Z^{(t, 0)}[e_{ii}]}_{2, \dots, (1+d^{(t-1)})} + \smashoperator[lr]{\sum_{v_j \in \Gamma_G(v_i)}} w_{ij} {Z^{(t, 0)}[e_{jj}]}_{2, \dots, (1+d^{(t-1)})}
	We now explicitly define parameter matrices for $S^{(t, 1)}$ and $S^{(t, 2)}$ s.t.\ this weighted sum is produced.
	Note that the weighted vertex neighborhood sum only requires scalar multiplication and vector addition, i.e.\ the $d^{(t-1)}$ vertex feature dimensions are mutually independent.
	W.l.o.g.\ this allows us to simplify notation by treating the vertex feature vectors as if they were scalars in the following definitions, i.e.\ we can assume $d^{(t-1)} = 1$ and $Z^{(t,0)}[e_{ii}] = (1, Z^{(t-1)}[v_i], w_{ii}) \in \mathbb{R}^{3}$.
	The layer $S^{(t, 1)}$ is defined by
		\begin{aligned}
			Z^{(t,1)}[e_{ij}] &= \textcolor{t_blue}{Z^{(t,0)}[e_{ij}] W_{\mathrm{L}}^{(t,1)}} + \smashoperator[lr]{\sum_{v_l \in {\Gamma_{G^1}(v_i) \cap \Gamma_{G^1}(v_j)}}} \textcolor{t_red}{\left( Z^{(t,0)}[e_{ij}] W_{\mathrm{F}}^{(t,1)}\right)} \odot \textcolor{t_darkgreen}{\left( \left( Z^{(t,0)}[e_{il}] + Z^{(t,0)}[e_{lj}] \right) W_{\Gamma}^{(t,1)} \right)} \\ % chktex 25
			&= \begin{cases}
				\textcolor{t_blue}{(1, 0, w_{ii}, 0)} + \textcolor{t_darkgreen}{\left(0, 0, 0, 2\textcolor{t_red}{w_{ii}} Z^{(t-1)}[v_i]\right)} + \smashoperator[lr]{\sum_{v_l \in \Gamma_{G^1}(v_i)}} \textcolor{t_darkgreen}{\left(0, w_{il} \textcolor{t_red}{Z^{(t-1)}[v_i]}, 0, 0\right)} & \text{if $i = j$} \\ % chktex 25
				\textcolor{t_blue}{(0, 0, w_{ij}, 0)} + \textcolor{t_darkgreen}{\left(0, 0, 0, \textcolor{t_red}{w_{ij}} Z^{(t-1)}[v_i]\right)} + \textcolor{t_darkgreen}{\left(0, 0, 0, \textcolor{t_red}{w_{ij}} Z^{(t-1)}[v_j]\right)} & \text{else} % chktex 25
			\end{cases} \\
			&= \begin{cases}
				\left(1, \mkern6mu\smashoperator[lr]{\sum_{v_l \in \Gamma_{G^1}(v_i)}} w_{il} Z^{(t-1)}[v_i], w_{ii}, 2 w_{ii} Z^{(t-1)}[v_i] \right) & \text{if $i = j$} \\
				\left(0, 0, w_{ij}, w_{ij} (Z^{(t-1)}[v_i] + Z^{(t-1)}[v_j])\right) & \text{else}
			\end{cases}
		\end{aligned}\\
		\text{with }
		W_{\mathrm{L}}^{(t,1)} \coloneqq \begin{pmatrix}
			\textcolor{t_blue}{1} & 0 & 0 & 0 \\ % chktex 25
			0 & 0 & 0 & 0 \\
			0 & 0 & \textcolor{t_blue}{1} & 0 % chktex 25
		\end{pmatrix},
		W_{\mathrm{F}}^{(t,1)} \coloneqq \begin{pmatrix}
			0 & 0 & 0 & 0 \\
			0 & \textcolor{t_red}{\frac{1}{2}} & 0 & 0 \\ % chktex 25
			0 & 0 & 0 & \textcolor{t_red}{1} % chktex 25
		\end{pmatrix},
		W_{\Gamma}^{(t,1)} \coloneqq \begin{pmatrix}
			0 & 0 & 0 & 0 \\
			0 & 0 & 0 & \textcolor{t_darkgreen}{1} \\ % chktex 25
			0 & \textcolor{t_darkgreen}{1} & 0 & 0 % chktex 25
		\end{pmatrix}
	The vertex neighborhood summation is completed via $S^{(t, 2)}$ which is defined by
		\begin{aligned}
			Z^{(t,2)}[e_{ij}] &= \begin{cases}
				\textcolor{t_blue}{\left(1, -\smashoperator[lr]{\sum_{v_l \in \Gamma_{G^1}(v_i)}} w_{il} Z^{(t-1)}[v_i], w_{ii} \right)} + \smashoperator[lr]{\sum_{v_l \in \Gamma_{G^1}(v_i)}} \textcolor{t_darkgreen}{\left(0, w_{il}\left(Z^{(t-1)}[v_i] + Z^{(t-1)}[v_l]\right), 0\right)} & \text{if $i = j$} \\ % chktex 25
				\textcolor{t_blue}{(0, 0, w_{ij})} & \text{else} % chktex 25
			\end{cases} \\
			&= \begin{cases}
				\left(1, w_{ii} Z^{(t-1)}[v_i] + \smashoperator[lr]{\sum_{v_l \in \Gamma_{G}(v_i)}} w_{il} Z^{(t-1)}[v_l], w_{ii}\right) & \text{if $i = j$} \\
				(0, 0, w_{ij}) & \text{else}
			\end{cases}
		\end{aligned}\\
		\text{with }
		W_{\mathrm{L}}^{(t,2)} \coloneqq \begin{pmatrix}
			\textcolor{t_blue}{1} & 0 & 0 \\ % chktex 25
			0 & \textcolor{t_blue}{-1} & 0 \\ % chktex 25
			0 & 0 & \textcolor{t_blue}{1} \\ % chktex 25
			0 & 0 & 0
		\end{pmatrix},\quad
		W_{\mathrm{F}}^{(t,2)} \coloneqq \begin{pmatrix}
			0 & \textcolor{t_red}{\frac{1}{2}} & 0 \\ % chktex 25
			0 & 0 & 0 \\
			0 & 0 & 0 \\
			0 & 0 & 0
		\end{pmatrix},\quad
		W_{\Gamma}^{(t,2)} \coloneqq \begin{pmatrix}
			0 & 0 & 0 \\
			0 & 0 & 0 \\
			0 & 0 & 0 \\
			0 & \textcolor{t_darkgreen}{1} & 0 % chktex 25
		\end{pmatrix}
	Using the two layers $S^{(t, 1)}$ and $S^{(t, 2)}$ that we just defined, the weighted vertex neighborhood sum for all $v_i \in \mathcal{V}_G$ is contained in $Z^{(t,2)}[e_{ii}]$.
	Additionally for all $e_{ij} \in \mathcal{E}_{G^1}$ the indicators ${Z^{(t,2)}[e_{ij}]}_1 = \mathbbm{1}[i = j]$ and the weights ${Z^{(t,2)}[e_{ij}]}_{d^{(t)} + 2} = w_{ij}$ are preserved.
	This means that the invariants~\ref{inv:ltd:wl2-simulation:indicator} and~\ref{inv:ltd:wl2-simulation:weight} are satisfied after $S^{(t, 2)}$.
	To complete the induction step it now remains to show that all three invariants hold after applying the layers $S^{(t, 2+1)}, \dots, S^{(t, 2+K)}$.
	As previously mentioned, a 2-\acs{wl} convolution layer is reduced to a fully connected layer if $W_{F}^{(t)} = \mathbf{0}$.
	Via the universal approximation theorem~\cite{Hornik1991} we can therefore use $S^{(t, 2+1)}, \dots, S^{(t, 2+K)}$ to simulate the $K$ layers of $\mathrm{\acs*{mlp}}^{(t)}$ without changing the first and last dimension of each feature vector to preserve the invariants~\ref{inv:ltd:wl2-simulation:indicator} and~\ref{inv:ltd:wl2-simulation:weight}.
	The resulting feature matrix $Z^{(t,2+K)}$ then satisfies all three invariants which completes the induction.

	Using invariants~\ref{inv:ltd:wl2-simulation:indicator} and~\ref{inv:ltd:wl2-simulation:feature} for $t = T$ we can therefore set
	\begin{align*}
		\varphi(Z_G^{(T,2+K)}[e_{ij}]) \coloneqq \begin{cases}
			{Z_G^{(T,2+K)}[e_{ij}]}_{2, \dots, (d^{T} + 1)} & \text{if ${Z_G^{(T,2+K)}[e_{ij}]}_1 = 1$} \\
			\texttt{nil} & \text{else}
		\end{cases}
		\text{.}
	\end{align*}
	By our previous definition of $\mathit{Pool}_2$, this in turn implies that $\mathit{Pool}_2(Z_G^{(T,2+K)}) = \mathit{Pool}_1(Z_G^{(T)}) \iff h_2(G) = h_1(G)$ which concludes the proof.
\end{proof}

### 2020-04-03 thesis/content/chapter-related-work.tex
started with 2-wl-gnn expressive power
completed proof that 2-wl-gnns are at least as powerful as 1-wl

\subsubsection{Spatial \acp{gcnn}}
Looking at the \ac{gcn} layer defined in \cref{eq:related:gcn-layer}, notice that, even though it is motivated by spectral graph theory, the spectrum $\lambda_G$ is not actually directly used.
This allows for a very different perspective on \acp{gcn}:
Instead of interpreting convolutions as applications of linear spectral filters, we can interpret them as vertex neighborhood aggregations, similar to 1-\acs{wl}.
From this perspective a \ac{gcn} can be seen as a so-called spatial \ac{gcnn} because it operates directly in the vertex domain.
To see why this is the case, we rewrite \cref{eq:related:gcn-layer} and compare it with the 1-\acs{wl} color refinement strategy (see \cfullref{defn:related:wl1-refine}):
\begin{alignat*}{2}
	\text{\ac*{gcn}:}\quad&& Z_G^{(t)}[v] &= \sigma\left(\left( {\mu(v)}^2 \textcolor{t_blue}{Z_G^{(t-1)}[v]} + \smashoperator[lr]{\sum_{u \in \Gamma_G(v)}} \mu(v) \mu(u) \textcolor{t_red}{Z_G^{(t-1)}[u]} \right) \Theta^{(t)} \right) \\ % chktex 25
	\text{1-\acs{wl}:}\quad&& \chi_{G,1}^{(t)}(v) &= h\left( \mkern60mu\textcolor{t_blue}{\chi_{G,1}^{(t-1)}(v)}, \ldblbrace \textcolor{t_red}{\chi_{G,1}^{(t-1)}(u)}\, |\, u \in \Gamma_{G}(v) \rdblbrace\right) % chktex 25
\end{alignat*}
Here $\mu(v) \coloneqq {\left(\left| \Gamma_G(v) \right| + 1\right)}^{-\frac{1}{2}}$ are normalization factors introduced by $\tilde{D}_G^{-\frac{1}{2}}$.
The comparison shows that \ac{gcn} convolutions can be understood as iterative refinements of continuous ``color vectors'' $Z_G^{(t)}[v] \in \mathbb{R}^{d^{(t)}}$.
Then the continuous analogue to the injective 1-\acs{wl} hash function $h: \mathcal{C}^{*} \to \mathcal{C}$ can be defined as $h_{\mathrm{\acs*{gcn}}}(z_0, z_1, \dots, z_m) = \sigma(\sum_{j=0}^m \mu(v_0) \mu(v_j) z_j \Theta^{(t)})$ which computes a single color vector from multiple color vectors $z_j$ where $v_j$ are the corresponding vertices.

\paragraph{A 1-\acs{wl} bound on \ac{gnn} power}
In the next step we will take a closer look at the relation between \acp{gcnn} that use vertex neighborhood convolution and 1-\acs{wl} by comparing their respective discriminative power.
\begin{prop}[see {\citet[lemma 2]{Xu2018}} for the proof]\label{prop:related:gcnn-wl1-limit}
	The discriminative power of a \ac{gcnn} that convolves vertex feature vectors via a convolution operator of the form $Z_G^{(t)}[v] = h^{(t)}\left(Z_G^{(t-1)}[v], \ldblbrace Z_G^{(t-1)}[u]\, |\, u \in \Gamma_G(v) \rdblbrace\right)$ is upper-bounded by that of 1-\acs{wl}, where $h^{(t)}: {(\mathbb{R}^{d^{(t-1)}})}^{*} \to \mathbb{R}^{d^{(t)}}$ is an arbitrary vertex neighborhood hashing function.
	Moreover, if $h^{(t)}$ is not injective, the \ac{gcnn} has a strictly lower discriminative power than 1-\acs{wl}.
\end{prop}
By \cref{prop:related:gcnn-wl1-limit} the \ac{gcn} architecture is strictly less powerful than 1-\acs{wl} due to the fact that $h_{\mathrm{\acs*{gcn}}}$ is not an injective hash function.
To overcome this limitation \citet{Xu2018} proposed an alternative vertex neighborhood convolution architecture called \ac{gin}:
\begin{align}
	Z_G^{(t)}[v] \coloneqq \mathrm{\acs*{mlp}}^{(t)}\left( Z_G^{(t-1)}[v] + \smashoperator[lr]{\sum_{u \in \Gamma_G(v)}} Z_G^{(t-1)}[u] \right)
\end{align}
By leaving out the normalization factors $\mu(v)$ and by using a \ac{mlp} instead of the single fully-connected layer $\sigma \circ \Theta^{(t)}$ the resulting hashing function $h_{\mathrm{\acs*{gin}}}(z_1, \dots, z_m) = \mathrm{\acs*{mlp}}^{(t)}(\sum_{j=1}^{m} z_j)$ becomes injective\footnote{
	Under the assumption that the set of possible input feature matrices $X \in \mathbb{R}^{n \times d^{(0)}}$ is countable.
	This is reasonable since real-world \ac{gcr} domains typically only have a finite set of possible vertex feature vectors, e.g\ one-hot label encodings.
}, i.e.\ it assigns a unique color vector to each multiset of color vectors.
By \cref{prop:related:gcnn-wl1-limit} this implies that \ac{gin} has the same discriminative power as 1-\ac{wl}.
As expected, \ac{gin} therefore fits training data better than non-injective \acp{gcnn} like \ac{gcn}.
Interestingly \ac{gin} additionally seems to generalize better on test data than non-injective methods~\cite{Xu2018}\cite{Errica2020}.
It is not yet fully understood why this is the case.

>>>>

\subsubsection{Spatial \acp{gcnn}}
Looking at the \ac{gcn} layer defined in \cref{eq:related:gcn-layer}, notice that, even though it is motivated by spectral graph theory, the spectrum $\lambda_G$ is not actually directly used.
This allows for a very different perspective on \acp{gcn}:
Instead of interpreting convolutions as applications of linear spectral filters, we can interpret them as vertex neighborhood aggregations, similar to 1-\acs{wl}.
From this perspective a \ac{gcn} can be seen as a so-called spatial \ac{gcnn} because it operates directly in the vertex domain.
To see why this is the case, we rewrite \cref{eq:related:gcn-layer} and compare it with the 1-\acs{wl} color refinement strategy (see \cfullref{defn:related:wl1-refine}):
\begin{equation}
	\begin{alignedat}{2}
		\text{\ac*{gcn}:}\quad&& Z_G^{(t)}[v] &= \sigma\left(\left( {\mu(v)}^2 \textcolor{t_blue}{Z_G^{(t-1)}[v]} + \smashoperator[lr]{\sum_{u \in \Gamma_G(v)}} \mu(v) \mu(u) \textcolor{t_red}{Z_G^{(t-1)}[u]} \right) \Theta^{(t)} \right) \\ % chktex 25
		\text{1-\acs{wl}:}\quad&& \chi_{G,1}^{(t)}(v) &= h\left( \mkern60mu\textcolor{t_blue}{\chi_{G,1}^{(t-1)}(v)}, \ldblbrace \textcolor{t_red}{\chi_{G,1}^{(t-1)}(u)}\, |\, u \in \Gamma_{G}(v) \rdblbrace\right) % chktex 25
	\end{alignedat}\label{eq:related:gcn-wl1}
\end{equation}
Here $\mu(v) \coloneqq {\left(\left| \Gamma_G(v) \right| + 1\right)}^{-\frac{1}{2}}$ are normalization factors introduced by $\tilde{D}_G^{-\frac{1}{2}}$.
The comparison shows that \ac{gcn} convolutions can be understood as iterative refinements of continuous ``color vectors'' $Z_G^{(t)}[v] \in \mathbb{R}^{d^{(t)}}$.
Then the continuous analogue to the injective 1-\acs{wl} hash function $h: \mathcal{C}^{*} \to \mathcal{C}$ can be defined as $h_{\mathrm{\acs*{gcn}}}(z_0, z_1, \dots, z_m) = \sigma(\sum_{j=0}^m \mu(v_0) \mu(v_j) z_j \Theta^{(t)})$ which computes a single color vector from multiple color vectors $z_j$ where $v_j$ are the corresponding vertices.

\paragraph{A 1-\acs{wl} bound on \ac{gnn} power}
In the next step we will take a closer look at the relation between \acp{gcnn} that use vertex neighborhood convolution and 1-\acs{wl} by comparing their respective discriminative power.
\begin{prop}[see {\citet[lemma 2]{Xu2018}} for the proof]\label{prop:related:gcnn-wl1-limit}
	The discriminative power of a \ac{gcnn} that convolves vertex feature vectors via a convolution operator of the form $Z_G^{(t)}[v] = h^{(t)}\left(Z_G^{(t-1)}[v], \ldblbrace Z_G^{(t-1)}[u]\, |\, u \in \Gamma_G(v) \rdblbrace\right)$ is upper-bounded by that of 1-\acs{wl}, where $h^{(t)}: {(\mathbb{R}^{d^{(t-1)}})}^{*} \to \mathbb{R}^{d^{(t)}}$ is an arbitrary vertex neighborhood hashing function.
	Moreover, if $h^{(t)}$ is not injective, the \ac{gcnn} has a strictly lower discriminative power than 1-\acs{wl}.
\end{prop}
By \cref{prop:related:gcnn-wl1-limit} the \ac{gcn} architecture is strictly less powerful than 1-\acs{wl} due to the fact that $h_{\mathrm{\acs*{gcn}}}$ is not an injective hash function.
To overcome this limitation \citet{Xu2018} proposed an alternative vertex neighborhood convolution architecture called \ac{gin}:
\begin{align}
	Z_G^{(t)}[v] \coloneqq \mathrm{\acs*{mlp}}^{(t)}\left( Z_G^{(t-1)}[v] + \smashoperator[lr]{\sum_{u \in \Gamma_G(v)}} Z_G^{(t-1)}[u] \right) \label{eq:related:gin-layer}
\end{align}
By leaving out the normalization factors $\mu(v)$ and by using a \ac{mlp} instead of the single fully-connected layer $\sigma \circ \Theta^{(t)}$ the resulting hashing function $h_{\mathrm{\acs*{gin}}}(z_1, \dots, z_m) = \mathrm{\acs*{mlp}}^{(t)}(\sum_{j=1}^{m} z_j)$ becomes injective\footnote{
	Under the assumption that the set of possible input feature matrices $X \in \mathbb{R}^{n \times d^{(0)}}$ is countable.
	This is reasonable since real-world \ac{gcr} domains typically only have a finite set of possible vertex feature vectors, e.g\ one-hot label encodings.
}, i.e.\ it assigns a unique color vector to each multiset of color vectors.
By \cref{prop:related:gcnn-wl1-limit} this implies that \ac{gin} has the same discriminative power as 1-\ac{wl}.
As expected, \ac{gin} therefore fits training data better than non-injective \acp{gcnn} like \ac{gcn}.
Interestingly \ac{gin} additionally seems to generalize better on test data than non-injective methods~\cite{Xu2018}\cite{Errica2020}.
It is not yet fully understood why this is the case.
